Bayesian exponential random graph modeling of KWL Burial network

### 2020-06-16 analysis/paper/paper.Rmd
work on method and reorganize paragraphs

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical model for expressing properties of network structures with a focus directly on possible ties between nodes by viewing them as random variables [@Robins2007]. In traditional statistical model for network, each predictor variable for forming a tie between two nodes is measure separately by the presence of a tie without considering interactions among ties [@Morris2008]. By contrast, statistical models assumes social networks as dependent variables that can represent the network dependencies between tie variables [@Snijders2011]. In a network generated by ERGMs, the formation of a tie is influenced by the presence or absence of other ties or attributes of nodes/ties. Predictor variables, called network statistics, such as reciprocity, transitivity, homophily, are direct functions of the ties themselves where the probability of occurrence of the configuration of ties can be specified and hypothesized [@Morris2008; @Robins2007]. With the development of Markov chain Monte Carlo (MCMC) algorithms that produce approximate maximum likelihood estimators and graph statistics for further specification of transitivity and heterogeneity, ERGMs can deal with complicated dependence patterns observed at one moment in time [@Snijders2006]. The advantage of networks specification under assumptions and contexts can be useful to explain archaeological network in a given time. @Amati2019 demonstrate the use of ERGMs to reconstruct networks consist of a set of sites in the Caribbean during the period AD 100–400. They found out the structure of network can be efficiently explained by multiple interdependent mechanisms instead of hypotheses of distance and cultural homophily exclusively. However, they also point out some limitations of application such as static outcome and sensitive to missing data. The two major limitations ERGMs are difficult to compute because their normalizing constant depending on model parameters is intractable. A Bayesian framework allows for parameter inference using Markov chain Monte Carlo (MCMC) strategies which avoid the need for computationally intensive calculations of the normalizing constants [@Caimo2011].

@@

Previous studies on burials suggest a presence of uneven distribution of prestige goods across burials. However, there is a debate over whether the differentiation between burials hints an economic inequality supporting an egalitarian society [@Hsieh2012], or political inequality indicating a stratified society [@Cheng2008]. @Cheng2008 interprets the unequal distribution of prestige materials, especially the gold-foil beads between burials, as evidence for hierarchy. However, @Hsieh2012 suggests a relatively egalitarian society based on the comparative analysis of frequency and proportion of all burial goods. She found that the burials with rich goods were usually associated with elders, which might indicate achieved, rather than inherited, status. The earliest record of the use of prestige goods in burials can be traced back around the European arrival according to historical records from the Spanish and the Chinese [@Borao2009; @LiandWu2006]. For example, the Spanish document by a Spanish Priest described that the Spanish soldier used agate beads to exchange materials from local indigenous people since they were valuable in their culture [@LiandWu2006]. In addition, there was a custom that people would bury imported materials with deceased to display their status and influence [@LiandWu2006]. Despite we observed the economic differentiation in burial goods and the high value of imported goods according to historical documents, the extent of economic differentiation and how it changed over time remain unclear. Analyzing and comparing burial data from different phases using network analysis approaches will give an insight into the degree and changes in social inequality. This is important to understand the social reactions of Kiwulan residents in the context of the foreign contact.

@@

<!--procedures-->

## data

We use burial data collected from the published excavation report, and original fieldwork note with detailed descriptions about context and stratigraphy. Given that burials in Chinese phase are less than previous two phases, we only compare these burials by their grave goods and consitions without further network analysis. Burial data provides information about the location, orientation, size, depth, age, gender, quantity of grave goods, and spatial relationship among them. We consider the effects of both exogenous and endogenous factors for meaningful explanation of the characteristics of burial network and tendency of burials correlate with each other. To be specific, our purpose is to understand the underlying mechanisms of the network structures and how their relationships have an influence on the accumulation of wealth between different foreign contact periods. We focus on connectivity, clusters, centralization, and homophily those properties of network that can reflect the social relations between individuals guided by social structures. Previous studies have shown that there is evidence of economic inequality between Kiwualn burials. However, the degree of inequality over time, and which variable decides the changes remain unclear.

We hypothesize that after the arrival of the European presence in the 17th century, a group of ambitious individuals competed with each other for trading with the Europeans due to the high value of trade goods in their society, which led to the accumulation of high quality trade goods. Social inequality gradually increased among Kiwulan society, which can be observed from the burials from a more equal distribution of grave goods to a more differentiation in goods accumulation. We hypothesize that the burial network will present less reciprocality, less transitivity, but higher popularity structure effects in burial network after the European presence.

We performed network analysis with Bayesian inference with the ergm package for simulation and model specification [@Hunter2008; @Morris2008], and then bergm package for Bayesian modeling using R programming [@Caimo2014]. We set the number of chains to eight. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10000. The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with a set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical way to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network.

Since the accumulation of imported prestige goods in burials is commonly observed, we assign the burials into four classes, high, medium, and none as an index of wealth. In addition to wealth, we also consider another two node variables, age and gender, serving as indicators for distinguishing ascribed or achieved status in a society [citation].

@@

it can accomodate binary network statistics to describe the weighted network topology [@Snijders2006]

some concepts:
- endogenous statistics that ties are modelled based on the existence of other ties
- exogenous statistics that ties are formed according to monadic or dyadic node attributes

We used ergm package in R for the model specification. Every term in an ERGM has an associated algorithm for computing its value for network [@Morris2008]. The terms we selected for network specification include:

1. edges: ties, a measure of density, equal to kstar(1) for undirected networks

2. gwesp (geometrically weighted edgewise shared partners): triad-based clustering based on a dyad-based configurations (not node-based) that counts the number of times that both nodes have a tie to a third node. It is a measure of the tendency for two actors who are tied to have x partners in common. In other words, a tendency for those with shared partners to become tied, or tendency of ties to cluster together.

3. gwnsp (geometrically weighted non-edgewise shared partner statistics): a function of the non-edgewise shared partner statistics. The edgewise and non-edgewise shared partner statistics reflect tendencies toward transitivity abd compared together would give a more comprehensive result.

4. gwdegree (geometrically weighted degree distribution): a parametric form that represents the frequency distribution for nodal degree, each node counts only once. The local configuration is degrees and stars that are equivalent representations for the distribution of node-based edge counts. A measure of the tendency of the number of edges into a node. This statistic captures the tendency towards centralisation in the degree distribution of the network.

5. nodematch: uniform homophily between nodes that refers to the similarity of connected nodes. A statistics that counts the number of edges between nodes sharing the same covariate value or characteristics. We use wealth index and age as covariants that reflects the similarity between nodes. Here we consider age, gender, and index of wealth as node attributes covariants.

>>>>

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical model for expressing properties of network structures with a focus directly on possible ties between nodes by viewing them as random variables [@Robins2007]. In traditional statistical model for network, each predictor variable for forming a tie between two nodes is measure separately by the presence of a tie without considering interactions among ties [@Morris2008]. By contrast, statistical models assumes social networks as dependent variables that can represent the network dependencies between tie variables [@Snijders2011]. In a network generated by ERGMs, the formation of a tie is influenced by the presence or absence of other ties or attributes of nodes/ties. Predictor variables, called network statistics, such as reciprocity, transitivity, homophily, are direct functions of the ties themselves where the probability of occurrence of the configuration of ties can be specified and hypothesized [@Morris2008; @Robins2007]. The advantage of networks specification under assumptions and contexts can be useful to explain archaeological network in a given time. For example, @Amati2019 demonstrate the use of ERGMs to reconstruct networks consist of a set of sites in the Caribbean during the period AD 100–400. They found the structure of network can be efficiently explained by multiple interdependent mechanisms instead of hypotheses of distance and cultural homophily exclusively. However, they also point out some limitations of application such as static outcome and sensitivity to missing data. Another limitation is that ERGMs are difficult to compute because their normalizing constant depending on model parameters is intractable. With the development of Markov chain Monte Carlo (MCMC) algorithms that produce approximate maximum likelihood estimators and graph statistics for further specification of transitivity and heterogeneity, ERGMs can deal with complicated dependence patterns observed at one moment in time [@Snijders2006]. A Bayesian framework allows for parameter inference using Markov chain Monte Carlo (MCMC) strategies which avoid the need for computationally intensive calculations of the normalizing constants [@Caimo2011].

@@

Previous studies on burials suggest a presence of uneven distribution of prestige goods across burials. However, there is a debate over whether the differentiation between burials hints an economic inequality supporting an egalitarian society [@Hsieh2012], or political inequality indicating a stratified society [@Cheng2008]. @Cheng2008 interprets the unequal distribution of prestige materials, especially the gold-foil beads between burials, as evidence for hierarchy. However, @Hsieh2012 suggests a relatively egalitarian society based on the comparative analysis of frequency and proportion of all burial goods. She found that the burials with rich goods were usually associated with elders, which might indicate achieved, rather than inherited, status. The earliest record of the use of prestige goods in burials can be traced back around the European arrival according to historical records from the Spanish and the Chinese [@Borao2009; @LiandWu2006]. For example, the Spanish document by a Spanish Priest described that the Spanish soldier used agate beads to exchange materials from local indigenous people since they were valuable in their culture [@LiandWu2006]. In addition, there was a custom that people would bury imported materials with deceased to display their status and influence [@LiandWu2006]. Despite we observed the economic differentiation in burial goods and the high value of imported goods according to historical documents, the degree of inequality over time, and which variable decides the changes remain unclear. Analyzing and comparing burial data from different phases using network analysis approaches will give an insight into the degree and changes in social inequality. This is important to understand the social reactions of Kiwulan residents in the context of the foreign contact.

@@

## data and hypoethesis

We analyze archaeological burial networks from a Iron Age site, Kiwulan, in northeastern Taiwan. We use burial data collected from the published excavation report, and original fieldwork note with detailed descriptions about context and stratigraphy. Given that burials in Chinese phase are less than previous two phases, we only compare the quantity and types of burial goods in burials in Chinese phase rather than network analysis. For the quantity of burial goods, we assigned the burials into three classes, high, medium, and low as an index of wealth as a variable to indicate the extent of the accumulation of wealth. In addition to wealth, we also consider another two social variables, age and gender, serving as indicators for distinguishing ascribed or achieved status in a society [citation].

The purpose of our analysis is to understand the underlying mechanisms of the network structures and how their relationships have an influence on the accumulation of wealth between different foreign contact periods. We hypothesize that after the arrival of the European presence in the 17th century, a group of ambitious individuals competed with each other for trading with the Europeans due to the high value of trade goods in their society, which led to the accumulation of high quality trade goods. Social inequality gradually increased among Kiwulan society, which can be observed from the burials from a more equal distribution of grave goods to a more differentiation in goods accumulation. Thus, we predict that the burial network will present less reciprocality, less transitivity, but higher popularity structure effects in burial network after the European presence compared to the network before the European presence.

@@

We use Bayesian reference on Exponential Random Graph Models (ERGMs) to conduct statistical network analysis to test what variables may have contributed to the economic inequality observed between burials over time under foreign contact conditions. ERGMs assume that each network tie is a random variable which is dependent to each other somehow through the shared attributes or a reciprocal relationship [@Robins2007]. The dependence of ties can be hypothesized by parameters correspond to configurations in the network, such as a single tie, or a transitive triad. A model is built up by a distribution of random graphs according to these configurations that are also called network statistics. According to the formation of ties, network statistics can be divided into two ways, endogenous statistics and exogenous statistics [@Amati2019]. In endogenous model, the dependence of ties is decided by the presence of other ties, while in exogenous model, the dependence of ties is determined by node attributes. We use both exogenous and endogenous statistics to understand the characteristics of burial network and tendency of burials correlate with each other. To describe the relationship linked by shared burial goods better, the network is weighted to capture the multiple effects [@Snijders2006]. The network variables we examine include connectivity, clusters, centralization, and homophily in a undirected relations to reflect the social relations between individuals guided by social structures. Table \@ref(tab:ergm-terms-table) lists the network statistics we used in the model construction for burials with archaeological interpretations. Every term in an ERGM has an associated algorithm for computing its value for network [@Morris2008].

```{r ergm-terms-table}
ergm_terms <- readxl::read_excel(here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
kableExtra::kable(ergm_terms,
             caption = "The parameters of exponential random graph models for a undirected network to corresponding with archaeological evidence.")
```

We conduct network analysis using R programming with the ergm package for simulation and model specification [@Hunter2008; @Morris2008], and then bergm package for Bayesian modeling [@Caimo2014].
We set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10000. The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with a set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical way to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network.
